\magnification=1200

\nopagenumbers

\vsize=10in
\voffset=-.3in

\input amstex
\loadmsbm

\def\ctln{\centerline}
\def\ssk{\smallskip}
\def\msk{\medskip}
\def\bsk{\bigskip}
\def\nidt{\noindent}
\def\del{\partial}
\def\bbr{{\Bbb R}}
\def\cla{{\Cal A}}
\def\clb{{\Cal B}}
\def\clc{{\Cal C}}
\def\ra{\rightarrow}
\def\lra{$\Leftrightarrow$}

\ctln{\bf Math 856 Problem Set 4}

\ssk

\ctln{Starred (*) problems to be handed in Friday, November 13}

\bsk


\item{\bf (*) 25.} Let $\pi:E\ra B$ be an orientable vector bundle, and
$f:X\ra B$ a smooth map. Show that the pullback $\pi^\prime:f^*E\ra X$
is an orientable bundle.

\msk

\item{\bf 26.} If $\xi$ is a vector bundle, show that the Whitney sun of $\xi$
with itself, $\xi\oplus \xi$, is orientable.

\msk

\item{\bf 27.} Let $X_1,\ldots ,X_k$ be linearly independent vector fields on a manifold
$M$ with Riemannian metric $\langle\cdot,\cdot\rangle$ . Show that the Gram-Schmidt
orthonormalization procedure can be applied to the vector fields at every point 
simultaneously, to give a collection of 
orthonormal vector fields $Y_1,\ldots ,Y_k$.


\item{\bf 28.} [Lee, p.289, problem 11-17] We defined the metric induced by a Riemannian
metric $\langle\cdot,\cdot\rangle_a$ to be the infimum of lengths of piecewise smooth
curves from $p$ to $q$. Show that this cannot be a \underbar{minimum}: For $M=\bbr^2\setminus(0,0)$
with the usual Riemannian metric from $\bbr^2$, show that no curve in $M$ from $(-1,0)$ to $(1,0)$
has length equal to the distance between the two points. [Also known as: I lied in class....]

\msk

\item{\bf 29.} (a) Show that an immersion from one $n$-manifold 
(with\underbar{out} boundary...) to another is an open map.

\ssk

\item{} (b) Show that if $M$ and $N$ are $n$-manifolds, $M$ is compact, $N$ is connected, and
$F:M\ra N$ is an immersion, then $F$ is onto.

\msk

\item{\bf (*) 30.} [Lee, p.171, problem 7-2] Show that under the 
quotient map $p:S^2\ra {\bbr}P^2$ given by $p(x)=\{x,-x\}$ that 
the map $f:S^2\ra{\bbr}^4$ given by $f(x,y,z)=(x^2-y^2,xy,xz,yz)$
descends to a smooth \underbar{embedding} $\overline{f}:{\bbr}P^2\ra {\bbr}^4$.

\msk

\item{\bf (*) 31.} [Lee, p.172, problem 7-9] Show that if $\pi:M\ra N$
is a submersion and $Y$ is a vector field on $N$, then there is a
vector field $X$ (called a {\it lift} of $X$) on $M$ that is 
$\pi$-related to $Y$. (Hint: partition of unity!)

\msk

\item{\bf 32.} If $S\subseteq M$ is a {\bf closed}, embedded submanifold, $U\supseteq S$
is an open neighborhood of $S$, and 
$f:S\ra \bbr$ is a smooth function, show that there is a smooth function
$F:M\ra \bbr$ with $F|_S=f$ and supp$(F)\subseteq U$.

\vfill
\end
