

\documentclass[12pt]{article}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{dutchcal}

\textheight=10in
\textwidth=6.5in
\voffset=-1in
\hoffset=-1in

\begin{document}

\def\ctln{\centerline}
\def\msk{\medskip}
\def\bsk{\bigskip}
\def\ssk{\smallskip}
\def\hsk{\hskip.3in}
\def\ra{\rightarrow}
\def\ubr{\underbar}
\def\dsp{\displaystyle}

\def\mt{{\mathcal T}}
\def\mb{{\mathcal B}}
\def\ms{{\mathcal S}}
\def\mu{{\mathcal U}}
\def\mv{{\mathcal V}}

\def\bbr{{\mathbb R}}
\def\bbz{{\mathbb Z}}
\def\bbq{{\mathbb Q}}
\def\bbn{{\mathbb N}}
\def\spc{$~$\hskip.15in$~$}

\def\sset{\subseteq}
\def\del{\partial}
\def\lra{$\Leftrightarrow$}
\def\bra{$\Rightarrow$}



%%\UseAMSsymbols

\ctln{\bf Math 325 Problem Set 3}

\msk

\ctln{Starred (*) problems are due Friday, September 14.}


\begin{description}

\item{10.} Show, by induction, that the (ordinary) triangle
inequality extends to show that for any $n\geq 2$ we have

\ssk

\ctln{$\big|\sum_{k=1}^n a_k\big| \leq \sum_{k=1}^n |a_k|$ .}

\msk

\item{(*) 11.} Show that the \ubr{maximum} of two numbers $x,y\in\bbr$ 
can be computed by 

\ssk

\ctln{$\dsp \textrm{max}(x,y) = {{x+y+|x-y|}\over{2}}$ .}

\ssk

[That is, if $x\leq y$ then $\dsp {{x+y+|x-y|}\over{2}} = y$, while if $y\leq x$ then it equals $x$.]

\ssk

Find a similar formula which gives the minimum of $x$ and $y$.


\msk

\item{12.} (Belding and Mitchell, p.35, \#4) Under what conditions on $a,b,c\in\bbr$ can we
conclude that $ab=cb$ implies that $a=c$ ? Prove your answer.

\msk

\item{13.} (Belding and Mitchell, p.3,5 \#7) If we have a subset $A\sset\bbr$ of real numbers, 
then $A$ satisfies all of the properties of a field so long as (a) 0 and 1 are elements of $A$ and
(b) whenever $a,b\in A$ we have $-a\in A$,  $a^{-1}\in A$, $a+b\in A$ and $ab\in A$ (where these 
are the negative and inverse \underbar{in} \underbar{the} \underbar{real} \underbar{line}); that is,
$A$ is {\it closed} under negation, inverse, addition, and multiplication.

\ssk

Use this to show that the set $\bbq[\sqrt{11}]=\{a+b\sqrt{11}\ :\ a,b\in\bbr\}$ satisfies all of the 
properties of a field.

\msk

\item{(*) 14.} (Belding and Mitchell, p.36, \#17) Use the triangle inequality to establish
that for every $x,y\in\bbr$ we have

\ssk

\item{(*)} (a) $|x|-|y|\leq |x-y|$
\hskip2in (*) (b) $|x|-|y|\leq |x+y|$

\ssk

\item{(*)} (d) $\Bigl||x|-|y|\Bigr| \leq |x-y|$

\msk

\item{(*) 15.} (a) Show that if $B\subseteq\bbr$ is {\it bounded}, and $A\subseteq B$, 
then $A$ is bounded.

\ssk

\item{(*)} (b) If $S\subseteq\bbr$, then we define the set $|S|$ as $|S|=\{|s|\ :\ s\in S\}$.
Show that if $S$ is bounded, then $|S|$ is bounded.

\msk

\item{16.} For each of the following sets, either show that it is bounded (and find bounds), or 
explain why it isn't. [You can appeal to results from calculus in your answers.]

\ssk

\item{\spc} (a) $\dsp A = \Big\{\sum_{k=1}^n {{1}\over{k}}\ :\ n\in\bbn\Big\}$
\hskip2in (b) $\dsp B = \Big\{\sum_{k=1}^n {{1}\over{2^k}}\ :\ n\in\bbn\Big\}$

\ssk

\item{\spc} (c) $\dsp C = \Big\{ {{\ln n}\over{n}}\ :\ n\in\bbn\Big\}$
\hskip2in (d) $\dsp D = \Big\{ {{2^n}\over{n^2}}\ :\ n\in\bbn\Big\}$


\end{description}
\vfill

\end{document}

