

\documentclass[12pt]{article}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{dutchcal}

\textheight=10in
\textwidth=6.5in
\voffset=-1in
\hoffset=-1in

\begin{document}

\def\ctln{\centerline}
\def\msk{\medskip}
\def\bsk{\bigskip}
\def\ssk{\smallskip}
\def\hsk{\hskip.3in}
\def\ra{\rightarrow}
\def\ubr{\underbar}
\def\dsp{\displaystyle}

\def\mt{{\mathcal T}}
\def\mb{{\mathcal B}}
\def\ms{{\mathcal S}}
\def\mu{{\mathcal U}}
\def\mv{{\mathcal V}}

\def\bbr{{\mathbb R}}
\def\bbz{{\mathbb Z}}
\def\bbq{{\mathbb Q}}
\def\bbn{{\mathbb N}}
\def\spc{$~$\hskip.15in$~$}

\def\sset{\subseteq}
\def\del{\partial}
\def\lra{$\Leftrightarrow$}
\def\bra{$\Rightarrow$}



%%\UseAMSsymbols

\ctln{\bf Math 325 Problem Set 7}

\msk

\ctln{Starred (*) problems are due Friday, October 19.}


\begin{description}


\item{(*) 38.} (Belding and Mitchell, p.89, \#10(a)) Show that if $a,b\in\bbr$, $a<b$, 
and $f:(a,b)\ra\bbr$ is {\it uniformly} continuous,
then $f$ is bounded: there are $M,N\in\bbr$ so that $M\leq f(x)\leq N$ for every $x\in (a,b)$.

\ssk

\item{\spc} Hint: the argument that we gave in class sort of works, but we can't `start' at
$a$ (it's not in the domain)!. `Start' in the middle, instead!

\msk

\item{39.} (Belding and Mitchell, p.88, \#5) Show that if $f,g:(a,b)\ra\bbr$ are both uniformly
continuous, then $f+g$ and $f-g$ (as functions from $(a,b)$ to $\bbr$) are also uniformly continuous.

\msk

\item{(*) 40.} (Belding and Mitchell, p., \#5, the other part) 
If $a,b\in\bbr$ and $f,g:(a,b)\ra\bbr$ are uniformly continuous, show that $fg:(a,b)\ra\bbr$ is uniformly
continuous. Show, by example, on the other hand that $f,g:\bbr\ra\bbr$ both uniformly continuous
does \ubr{not} imply that $fg:\bbr\ra\bbr$ is uniformly continuous.

\ssk

\item{\spc} [Hint: for the first, try to follow the `usual' continuity argument for products; problem \#38 will
help! for the second, you don't need to get very creative...]

\msk

\item{41.} (Belding and Mitchell, p.89, \#14) Let $f$ and $g$ be uniformly continuous on $\bbr$. Prove that their 
composition $f\circ g$ is also uniformly continuous on $\bbr$ .
 
\msk

\item{42.} (Belding and Mitchell, p.89, \#12) A function $f:D\ra\bbr$ is called {\it Lipschitz} (with Lipschitz
constant $M\in\bbr$ if for every $x,y\in D$ we have $|f(x)-f(y)|\leq M\cdot|x-y|$. Show that a Lipschitz function 
is uniformly continuous on $D$. Use this to show that the function $f:\bbr\ra\bbr$ defined by

\ssk

\ctln{$\dsp f(x)={{1}\over{1+|x|}}$}

\ssk

\item{\spc} is uniformly continuous on $\bbr$ .

\msk

\item{43.} Suppose that $f:[a,b]\ra\bbr$ is continuous, and define a new function $g:[a,b]\ra\bbr$ by
$g(x)={\textrm lub}\{f(t)\ :\ a\leq t\leq x\}$. Show that $g$ is also continuous!

\ssk

\item{\spc} [Hint: that least upper bound is (always) achieved!]

\msk

\item{(*) 44.} A function $f:\bbr\ra\bbr$ is called {\it periodic} if there is a $P>0$ so that $f(x+P)=f(x)$ for every $x\in\bbr$.
Show that if $f$ is periodic and continuous, then $f$ is uniformly continuous. 

\ssk

\item{\spc} [Hint: you can always insist that your $\delta$'s be smaller than some particular positive number (why?).]

\msk


\end{description}
\vfill

\end{document}

