\magnification=1200

\nopagenumbers

\vsize=10in
\voffset=-.3in

\input amstex
\loadmsbm

\def\ctln{\centerline}
\def\ssk{\smallskip}
\def\msk{\medskip}
\def\bsk{\bigskip}
\def\nidt{\noindent}
\def\del{\partial}
\def\bbr{{\Bbb R}}
\def\cla{{\Cal A}}
\def\clb{{\Cal B}}
\def\clc{{\Cal C}}
\def\ra{\rightarrow}
\def\lra{$\Leftrightarrow$}

\ctln{\bf Math 856 Problem Set 4}

\ssk

\ctln{Starred (*) problems to be handed in any time. Target date: Thursday, December 1}

\bsk

\item{\bf 21.} [Lee, p.101, problem 5-8] Let $p:E\ra M$ be a smooth $n$-dimensional 
vector bundle and 
$X_1,\ldots ,X_k$ be linearly independent smooth sections of $E$ defined over an open 
subset $U\subseteq M$.
Show that for every $a\in U$ there is a neighborhood $V$ of $a$ and smooth sections
$Y_{k+1},\ldots ,Y_n$ defined over $V$ so that $(X_1,\ldots ,X_k,Y_{k+1},\ldots ,Y_n)$
forms a local frame for $E$ over $U\cap V$. 

\ssk

\item{} (Hint: if $v_1,\ldots ,v_n$ form a basis for $\bbr^n$, then why is it that if you wiggle 
the first $k$ vectors a little bit, you still have a basis?)
\msk

\msk

\item{\bf (*) 22.} Show that if $M$ is a smooth manifold with boundary, then $\del M\subseteq M$
is an embedded submanifold. [So it has a unique smooth structure making the inclusion 
$\iota:\del M\rightarrow M$ a smooth embedding.]

\msk

\item{\bf 23.} (a) Show that an immersion from one $n$-manifold to another is an open map.

\ssk

\item{} (b) Show that if $M$ and $N$ are $n$-manifolds, $M$ is compact, $N$ is connected, and
$F:M\ra N$ is an immersion, then $F$ is onto.

\msk

\item{\bf (*) 24.} If $S\subseteq M$ is a {\bf closed}, embedded submanifold, $U\supseteq S$
is an open neighborhood of $S$, and 
$f:S\ra \bbr$ is a smooth function, show that there is a smooth function
$F:M\ra \bbr$ with $F|_S=f$ and supp$(F)\subseteq U$.

\msk

\item{\bf 25.} Let $X_1,\ldots ,X_k$ be linearly independent vector fields on a manifold
$M$ with Riemannian metric $\langle\cdot,\cdot\rangle$ . Show that the Gram-Schmidt
procedure can be applied to the vector fields all at once, to give a collection of 
orthonormal vector fields $Y_1,\ldots ,Y_k$.

\msk

\item{\bf 26.} [Lee, p.289, problem 11-17] We defined the metric induced by a Riemannian
metric $\langle\cdot,\cdot\rangle_a$ to be the infimum of lengths of piecewise smooth
curves from $p$ to $q$. Show that this cannot be a \underbar{minimum}: For $M=\bbr^2\setminus(0,0)$
with the usual Riemannian metric from $\bbr^2$, show that no curve in $M$ from $(-1,0)$ to $(1,0)$
has length equal to the distance between the two points. 

\msk

\item{\bf (*) 27.} [Lee, p.286, problem 11-3] Let $V$ and $W$ be finite-dimensional vector spaces over $\bbr$.
Show that there is a canonical (i.e., basis-independent) isomorphism between $V^*\otimes W$ and
the space $Hom(V,W)$ of linear maps from $V$ to $W$.


\vfill
\end
