\magnification=1200

\nopagenumbers

\input amstex
\loadmsbm

\def\ctln{\centerline}
\def\ssk{\smallskip}
\def\msk{\medskip}
\def\bsk{\bigskip}
\def\nidt{\noindent}
\def\del{\partial}
\def\bbr{{\Bbb R}}

\def\ra{\rightarrow}
\def\lra{$\Leftrightarrow$}



\ctln{\bf Math 856 Homework 2}

\ssk

\ctln{Starred (*) problems to be handed in Thursday, September 21}

\bsk

\nidt {\bf (*) 9:} Let ${\Bbb H}^n=\{(x_1,\ldots,x_n)\in\bbr^n : x_n\geq 0\}$. Suppose that 
$f:{\Bbb H}^n\rightarrow \bbr$ is a function for which every $x\in {\Bbb H}^n$ has an open 
neighborhood $U_x$ of $x$ such that $f$ extends to a $C^\infty$ function on $U_x$. Show that $f$ extends
to a $C^\infty$ function on a neighborhood of ${\Bbb H}^n$ in $\bbr^n$. 
(Short version: show that if $f$ is locally smooth, then $f$ is smooth. A well-chosen partition
of unity might help?)

\msk

\nidt{\bf 10:} We know that if $C,D\subseteq M$ are disjoint closed sets of the smooth manifold $M$,
then there exists a smooth function $f:M\ra [0,1]$ with $C\subseteq f^{-1}(0)$ and $D\subseteq f^{-1}(1)$.
But we can if fact make these containments {\it equalities}:

\ssk

(a) Show that it suffices to build a smooth function $g:M\ra [0,1]$ with $C=g^{-1}(0)$.

\ssk

(b) Build a countable cover $\{U_i\}$ of $M\setminus C$ by open sets of the form $h_i^{-1}(B(x_i,1)$ for
a collection of coordinate charts $h_i=(x^1,\ldots,x^n)$ with image containing $B(x_i,2)$. Build $C^\infty$ functions $g_i:M\ra \bbr$ which are $>0$
in $U_i$ and $=0$ on $M\setminus U_i$. Note that $\overline{U_i}$ is compact; for each $i$, let

\ssk

\ctln{$\displaystyle \alpha_i = \sup_{x\in \overline{U_i};j\leq i;m\leq i;k_1,\ldots 
k_m\leq n}\{ {{\del^m g_j}\over{\del x^{k_1}\cdots \del x^{k_m}}}(x)\}$ .}

\ssk

\noindent Show that the function $g=\sum g_i/(\alpha_i 2^i)$ is $C^\infty$ and $C=g^{-1}(0)$ .

\msk

\nidt {\bf 11:} Giving $M_1\times M_2$ the product smooth structure, show that $f:N\ra M_1\times M_2$
is smooth \lra\ the maps $p_1\circ f:N\ra M_1$, $p_2\circ f:N\ra M_2$ are smooth, where
$p_1,p_2$ are the projections onto the first and second factors, respectively. Show, moreover, that
the product smooth structure is the only smooth structure with this property.


\msk

\nidt {\bf 12:} If $M,N$ are smooth manifolds, show that $M\times N$ is diffeomorphic to $N\times M$.

\msk

\nidt {\bf (*) 13:} [Lee, problem 2.6] For $M$ a (smooth) manifold, let $C(M)$ denote the
set of continuous functions from $M$ to $\bbr$, thought of as an algebra (i.e., a ring and a vector
space over $\bbr$) with scalar multiplication by $\bbr$, and pointwise addition and multiplication.
Let $C^\infty(M)$ be the subalgebra of smooth functions. If $F:M\ra N$ is continuous,
let $F^*:C(N)\ra C(M)$ be given by $F^*(f)=f\circ F$.

\ssk

(a) Show that $F^*$ is a linear map.

\ssk

(b) Show that $F$ is smooth \lra\ $F^*(C^\infty(N))\subseteq C^\infty(M)$ .

\ssk

(c) Suppose $F$ is a homeomorphism. Show that $F$ is a diffeomorphism

\noindent \lra\ $F^*:C^\infty(N)\ra C^\infty(M)$ is an isomorphism.




\vfill
\end
