
\magnification=1200

\parindent=0pt
\nopagenumbers

\def\ctln{\centerline}
\def\msk{\medskip}
\def\bsk{\bigskip}
\def\hsk{\hskip10pt}

\ctln{\bf Topics for the Math 970/971 Qualifying Exam}

\bsk

\hsk {\bf Point-set Topology}

\msk

{\bf Toplogical spaces, continuous functions:} \hsk
topology, basis, subbasis, subspace topology, product topology,
quotient topology, closed sets, closure, limit points, continuous functions,
Pasting Lemma.

\msk

{\bf Homeomorphism invariants:} \hsk
separation properties ($T_0$, $T_1$, Hausdorff,  regular, normal),
countability properties, connectedness, path connectedness, 
compactness; local connectedness, local path connectedness,
metrizability. Applications.

\msk

{\bf Occasional topics:} \hsk
Tychonoff Theorem, Limit point compactness, 
sequential compactness, completeness, Urysohn's Lemma, 
Urysohn Metrization Theorem, Tietze Extension Theorem, manifolds.

\bsk

\hsk {\bf Homotopy and Homology}

\msk

{\bf Fundamental groups and covering spaces:} \hsk
homotopic maps, homotopy equivalence, cell complexes, 
fundamental group, induced homomorphisms, 
free groups, group presentations, Seifert - van Kampen Theorem,
covering spaces, lifting theorems, universal covering, 
deck transformations, normal coverings, Galois correspondence. 
Applications to group theory.

\msk

{\bf Homology theory:} \hsk
simplicial homology, singular homology, 
induced homomorphism, homotopy invariance, 
exact sequences, long exact homology sequence,
relative homology, homology with coefficients,
Mayer-Vietoris sequence, excision, 
Euler characteristic. Applications.

\msk

{\bf Occasional topics:} \hsk
Classification of surfaces, cellular homology, 
higher homotopy groups,
Hurewicz homomorphism.




\vfill
\end