

\documentclass[12pt]{article}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{dutchcal}

\textheight=10in
\textwidth=6.5in
\voffset=-1in
\hoffset=-1in

\begin{document}

\def\ctln{\centerline}
\def\msk{\medskip}
\def\bsk{\bigskip}
\def\ssk{\smallskip}
\def\ra{\rightarrow}
\def\ubr{\underbar}
\def\sset{\subseteq}
\def\smin{\setminus}

\def\mt{{\mathcal T}}
\def\mb{{\mathcal B}}
\def\ms{{\mathcal S}}
\def\mu{{\mathcal U}}
\def\mv{{\mathcal V}}
\def\mc{{\mathcal C}}

\def\mtp{{\mathcal T}^\prime}
\def\mbp{{\mathcal B}^\prime}
\def\mcp{{\mathcal C}^\prime}

\def\bbr{{\mathbb R}}
\def\bbz{{\mathbb Z}}
\def\spc{$~$\hskip.15in$~$}





%%\UseAMSsymbols

\ctln{\bf Math 871 Problem Set 10}

\msk

Starred (*) problems are due Thursday, December 4.


\begin{description}

\item{65.} [Munkres, p.330, Problem \#3(c)(d)] Show that if $X$ is a contractible 
space, then any two maps $f,g:Y\ra X$ are homotopic. Show in addition that if
$Y$ is path connected, then any two maps $f,g:X\ra Y$ are homotopic.

\msk

\item{(*)} 66. The {\it cone} $cX$ on a space $X$ is the quotient space $X\times I/\sim$ where
$(x,t)\sim(y,s)$ if $(x,t)=(y,s)$ or $t=s=1$. (That is, we crush $X\times\{1\}$ to a point.)
Show that for any space $X$, $cX$ is a contractible space.

\ssk

\item{\spc} [One approach: $cX$ can be viewed as a certain mapping cylinder.]

\msk

\item{(*)} 67. [Hatcher, p.18, Problem \#3(c)] Show that if maps $f,g:X\ra Y$ are
homotopic and $f$ is a homotopy equivalence, then $g$ is also a homotopy equivalence.

\msk

\item{68.} [Hatcher, p.19, Problem \#11] Show that if $f:X\ra Y$ is a map so that there
are maps $g,h:Y\ra X$ with $f\circ g\simeq\textrm{Id}_Y$ and $h\circ f\simeq\textrm{Id}_X$,
then $f$ is a homotopy equivalence.

\msk

\item{69.} Show tht if $f,g:X\ra Y$ and $h,k:Z\ra W$ satisfy $f\simeq g$ and $h\simeq k$,
then $f\times h,g\times k:X\times Z\ra Y\times W$ are homotopic.

\msk

\item{(*)} 70. Show that if $X$ is Hausdorff and $A\sset X$ is a retract of $X$, then $A$ is
a closed subset of $X$. 

\ssk

\item{\spc} [Hint: Problem \#40 of Problem Set 6 is useful here... find functions $f,g:X\ra X$ so that
$A$ is the set of points where the functions agree!]

\msk

\item{71.} [Munkres, p.186, Problem \#11a] Show that if $p:X\ra Y$ is a quotient
map, and $Z$ is a locally compact Hausdorff space, then 
$p\times\textrm{Id}_Z:X\times Z\ra Y\times Z$ is also a quotient map.

\ssk

\item{\spc} [See the very detailed hints in Munkres' text!]

\msk

\end{description}
\vfill

\end{document}



\item{72.} 


