

\documentclass[12]{article}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{dutchcal}
\usepackage{amsmath}

\textheight=10in
\textwidth=6.5in
\voffset=-1in
\hoffset=-1in

\begin{document}

\def\ctln{\centerline}
\def\msk{\medskip}
\def\bsk{\bigskip}
\def\ssk{\smallskip}
\def\ra{\rightarrow}
\def\ubr{\underbar}
\def\sset{\subseteq}

\def\mt{{\mathcal T}}
\def\mb{{\mathcal B}}
\def\ms{{\mathcal S}}
\def\mu{{\mathcal U}}
\def\mv{{\mathcal V}}

\def\bbr{{\mathbb R}}
\def\bbz{{\mathbb Z}}
\def\spc{$~$\hskip.15in$~$}





%%\UseAMSsymbols

\ctln{\bf Math 871 Problem Set 7}

\msk

Starred (*) problems are due Thursday, October 16.


\begin{description}

\item{42.} Show that if $h:(X\mt)\ra (Y,\mt^\prime)$ is a homeomorphism,
$A\subseteq X$, and $h(A)=B\subseteq Y$, then \hfill

\item{\spc} $h|_A:(A,\mt_A)\ra (B,\mt^\prime_B)$
is also a homeomorphism.



\msk

\item{43.} [Munkres, p.144, Problem \#2]

\item{(*)} (a) Show that if $p:(X,\mt)\ra(Y,\mt^\prime)$ is continuous,
and there is a continuous map $f:(Y,\mt^\prime)\ra(X,\mt)$ with
$p\circ f = Id_Y$, then $p$ is a quotient map.

\ssk

\item{(*)} (b) For $A\subseteq X$, a {\it retraction} 
of $X$ onto $A$ is a continuous map
$r:X\ra A$ with $r|_A=Id_A$. Show that a retraction is a quotient map.


\msk

\item{44.} [Munkres, p.152, Problem \#1] Show that if $\mt\subseteq \mt^\prime$ are topologies on
$X$ and $(X,\mt^\prime)$ is connected, then so is $(X,\mt)$.

\msk

\item{45.} Find an example of a topological space $(X,\mt)$ 
and $A\sset X$ so that $A$ is connected, but int$(A)$ isn't.

(Note: A result from class says that this \u{can't} be done in $\bbr$.
But it can be done in $\bbr^2$ ... or elsewhere!)


\msk

\item{(*)} 46. [Munkres, p.152, Problem \#9] Suppose that 
$A\subsetneq X$ and $B\subsetneq Y$ are {\it proper} subsets of $X$ and $Y$.
Show that if $(X\mt)$ and $(Y,\mt^\prime)$ are both connected, then 
$(X\times Y)\setminus(A\times B)$ is a connected subset of $X\times Y$.

\msk

\item{47.} Give an example of a space $X$ and subset $A\subseteq X$ where 
int$(A)$ and cl$(A)$ are both connected, but $A$ is not.

\msk

\item{(*)} 48. [Munkres, p.152, Problem \#11]
Suppose that $p:X\ra Y$ is a quotient map, $Y$ is connected, and 
for every $y\in Y$ the set $p^{-1}(\{y\})$ is a connected subset of $X$.
Show that $X$ is connected.

\ssk

\item{\spc} [Hint: start with a potential separation for $X$; show that their
images would separate $Y$.]


\end{description}
\vfill

\end{document}




