

\documentclass[12pt]{article}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{dutchcal}

\textheight=10in
\textwidth=6.5in
\voffset=-1in
\hoffset=-1in

\begin{document}

\def\ctln{\centerline}
\def\msk{\medskip}
\def\bsk{\bigskip}
\def\ssk{\smallskip}
\def\ra{\rightarrow}
\def\ubr{\underbar}
\def\sset{\subseteq}
\def\smin{\setminus}

\def\mt{{\mathcal T}}
\def\mb{{\mathcal B}}
\def\ms{{\mathcal S}}
\def\mu{{\mathcal U}}
\def\mv{{\mathcal V}}

\def\mtp{{\mathcal T}^\prime}
\def\mbp{{\mathcal B}^\prime}

\def\bbr{{\mathbb R}}
\def\bbz{{\mathbb Z}}
\def\spc{$~$\hskip.15in$~$}





%%\UseAMSsymbols

\ctln{\bf Math 871 Problem Set 8}

\msk

Starred (*) problems are due Thursday, November 6.


\begin{description}



\item{49.} Show that if $X_\alpha, \alpha\in I$ are all path-connected, then so is
$\prod_{\alpha\in I} X_\alpha$, if we use the product topology.

\msk

\item{(*)} 50. Show that if $A_\alpha\subseteq X, \alpha\in I$ are all path-connected, and
for some $\alpha_0\in I$ we have $\displaystyle A_\alpha\cap A_{\alpha_0}\neq\emptyset$ for all
$\alpha\in I$, then $\displaystyle\bigcup_{\alpha\in I}A_\alpha$ is path-connected.

\msk

\item{51.} Show that if $\mt$ and $\mtp$ are toplogies on $X$,
$(X,\mtp)$ is compact, and $\mt\sset\mtp$, then
$(X,\mt)$ is compact.


\msk

\item{52.} Give an example of a space $(X,\mt)$ and subsets $A,B\sset X$ so that
$A$ and $B$ are compact but $A\cap B$ is not.

\ssk

\item{\spc} (Note: your space $X$ cannot be Hausdorff....)


\msk

\item{(*)} 53. Let $X$ = $\bbr$ with the infinite ray topology 

\ssk

\ctln{$\mt$ = $\{(a,\infty) : a\in \bbr\}\cup \{\emptyset,\bbr\}$}

\ssk

\item{\spc} Show that $A$ = $\{0\}$ is a compact subset of $X$, but its closure $\overline{A}$ \ubr{isn't}.


\msk

\item{54.} [Munkres, p.171, Problem \#5] 
Show that if $(X,\mt)$ is a Hausdorff space and $A,B\sset X$ are disjoint compact
subsets of $X$, then there are subsets $U,V\in \mt$ so that $A\sset U$, $B\sset V$,
and $U\cap V =\emptyset$ .


\msk

\item{55.} [Munkres, p.171, Problem \#6] Show that if $f:(X,\mt)\ra (Y,\mtp)$ is continuous and 
surjective, $(X,\mt)$ is compact, and $(Y,\mtp)$ is Hausdorff, then $f$ is a closed map (and
therefore a quotient map).


\msk

\item{(*)} 56. Let $(X,\mt)$ be a Hausdorff space and let 

\ssk

\ctln{$\mtp = \{U\sset X\ :\ X\smin U\sset X\ \textrm{is compact}\}\cup\{\emptyset\}$.}

\item{\spc} Show that
$\mtp$ is a topology on $X$, and is coarser than $\mt$. Show that, in general, 
the two topologies need not be equal. Do $(X,\mt)$ and $(X,\mtp)$ have the same compact
subsets? [Note that problem \#51 says something about this...]

\ssk

\item{\spc} [This problem (except for the last part) was lifted from an old qualifying exam...] 

\end{description}
\vfill

\end{document}




