%\baselineskip=18pt plus 2pt

\magnification=1200
\overfullrule=0pt

\voffset=-.5in
\vsize=10in

\parindent=50pt

\def\ni{\noindent}
\def\ctln{\centerline}
\def\msk{\medskip}
\def\ssk{\smallskip}
\def\bsk{\bigskip}

\def\iit{\itemitem}
\def\htp{\hskip10pt}
\def\vtp{\vskip.02in}
\def\hsk{\hskip.2in}
%\input shorthand

\nopagenumbers

\ctln{\bf Math 856 Introduction to Smooth Manifolds}

%\vtp

\ctln{\bf Section 001}

\smallskip

\ni{\bf Lecture:} MWF 2:30-3:20 \htp (currently) Nebraska Hall (NH) W193

\msk

\ni{\bf Instructor:} Mark Brittenham

%\smallskip

\ni{\bf Office:} Avery Hall (AvH) 317
%\smallskip

\ni{\bf Telephone:} (47)2-7222

%\ssk

\ni{\bf E-mail:} mbrittenham2@math.unl.edu

\ni{\bf WWW:} http://www.math.unl.edu/$\sim$mbrittenham2/

\ni{\bf WWW pages for this class:} http://www.math.unl.edu/$\sim$mbrittenham2/classwk/856f11/

\ssk

\ni (There you will find copies of every handout from class, problem 
sets, class notes, and other items of interest.)

\ni{\bf Class lockbox:} http://www.math.unl.edu/$\sim$mbrittenham2/classwk/856f11/lockbox/

\ssk

\ni (There you will find things (articles, solutions) that shouldn't be given public access.
Username and password will be provided shortly after you read this.)

\smallskip

\ni{\bf Office Hours:} (tentatively) Mo 1:00-2:00, Tu 10:30-11:30, 
W 11:00 - 12:00, and whenever you can find me in my office and I'm not 
horrendously busy. You are also quite welcome to make an appointment
for any other time; this is easiest to arrange just before or 
after class, or via email.

\ssk

\ni{\bf Text:} {\it Introduction to Smooth Manifolds}, 
by John M. Lee (1st edition, Springer Verlag).

\msk

\ni This course, as its name is meant to imply, is intended to introduce you to 
the theory, techniques, and applications of smooth manifolds; broadly speaking,
this subject is known as differential 
topology. In basic outline we will follow the text; the specific topics covered
will depend partly on student interest. 

\msk

\ni{\bf Homework} will be assigned approximately every two weeks, and collected
one week after it is assigned. Somewhat more problems will be assigned than are collected;
those to be collected will be announced when the problems are assigned. The problems 
collected will be graded and returned. These grades will form the (only) basis 
for your final course grade. There will be no midterm or final exam in this class.

\msk

\ni{\bf Departmental Grading Appeals Policy:} Students who believe their
academic evaluation has been prejudiced or capricious have recourse for appeals 
to (in order) the instructor, the departmental chair, the departmental appeals 
committee, and 
the college appeals committee.

\msk

\ctln{\bf Some important academic dates}

\ssk

{\bf Aug. 22 (Mon.)} First day of classes.

{\bf Sept. 2 (Fri.)} Last day to withdraw from a course without a {\bf `W'}.

{\bf Sept. 5 (Mon.)} Labor Day - no classes (not that it matters for this class).

{\bf Oct. 14 (Fri.)} Last day to change to or from P/NP.

{\bf Oct. 17-18 (Mon.-Tue.)} Fall break - no classes. 

\item{} \hskip.2in (But with an AMS meeting on campus the
weekend before, 

\item{} \hskip.2in we won't have gotten far away?)

{\bf Nov. 11 (Fri.)} Last day to withdraw from a course.

{\bf Nov. 23 (Wed.)} Student holiday - no classes.

{\bf Nov. 24-27 (Thu.-Sun.)} Thanksgiving Vacation - no classes.

{\bf Dec. 10 (Sat.)} Last day of classes.

{\bf Dec. 12-16 (Mon.-Fri.)} Final exam week (not that we care for this class).

{\bf Dec. 16 (Fri.)} Fall semester graduate commencement.

\vfill

\end

\vfill\eject
